---
title: "Final Project (Technical Report)"
author: "Acree Tejada and Anwar Gafar"
format: pdf
editor: visual
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message=FALSE, warning=FALSE)
```

```{r}
#| include: false
#| message: false
#| warning: false

# to install additional packages, remove # and run following (you need to install a package only once)
# install.packages("package name")
```

```{r}
#| include: false
#| message: false
#| warning: false

library(tidyverse)
library(GGally)
library(caret)
library(recipes)
library(rsample)
library(rpart)
library(rpart.plot)
library(ranger)
library(e1071)
library(gbm)
library(kernlab)
library(pROC)
library(vip)
library(factoextra)
library(glmnet)

```

#### load datasets

```{r}

hotel_bookings_train <- read_csv("hotel_bookings_train.csv")
hotel_bookings_test <- read_csv("hotel_bookings_test.csv")

# Convert is_canceled to a factor with "yes" as the positive class
hotel_bookings_train <- hotel_bookings_train |>
  mutate(is_canceled = factor(tolower(is_canceled), levels = c("yes", "no")))


```

```{r}
### Feature Engineering

# Single Tree Recipe
# rpart handles factors so no dummy encoding or normalization needed.
# step_other reduce noise from sparse categories.
hotel_recipe_tree <- recipe(is_canceled ~ ., data = hotel_bookings_train) |>
  step_rm(id) |>
  step_mutate(
    total_nights           = stays_in_weekend_nights + stays_in_week_nights,    # total stay length
    total_guests           = adults + children + babies,                        # party size
    has_children           = as.integer(children > 0 | babies > 0),            # family flag
    room_type_changed      = as.integer(as.character(reserved_room_type) !=    # as.character() avoids
                                        as.character(assigned_room_type)),      # factor level mismatch error
    prior_cancel_rate      = previous_cancellations /                           # historical cancel behavior
                             (previous_cancellations + previous_bookings_not_canceled + 1),
    no_deposit             = as.integer(deposit_type == "No Deposit"),          # no financial commitment = higher cancel risk
    change_rate            = booking_changes / (lead_time + 1),                 # booking stability over time
    no_deposit_x_lead      = no_deposit * lead_time,                           # interaction: no deposit + long lead time
    risky_booking          = no_deposit * prior_cancel_rate,                   # interaction: no deposit + cancel history
    no_deposit_no_requests = as.integer(no_deposit == 1 &                      # no deposit + no special requests = low engagement
                                        total_of_special_requests == 0)
  ) |>
  step_impute_median(all_numeric_predictors()) |>   # median imputation is robust to outliers
  step_impute_mode(all_nominal_predictors()) |>     # mode imputation for categorical NAs
  step_novel(all_nominal_predictors()) |>           # handles unseen factor levels in test data
  step_string2factor(all_nominal_predictors()) |>   # convert characters to factors for rpart
  step_other(all_nominal_predictors(), threshold = 0.01)

hotel_recipe_tree


# Logistic Regression Recipe
# glmnet requires step_dummy and step_normalize
hotel_recipe_glm <- recipe(is_canceled ~ ., data = hotel_bookings_train) |>
  step_rm(id) |>
  step_mutate(
    total_nights           = stays_in_weekend_nights + stays_in_week_nights,
    total_guests           = adults + children + babies,
    has_children           = as.integer(children > 0 | babies > 0),
    room_type_changed      = as.integer(as.character(reserved_room_type) !=
                                        as.character(assigned_room_type)),
    prior_cancel_rate      = previous_cancellations /
                             (previous_cancellations + previous_bookings_not_canceled + 1),
    no_deposit             = as.integer(deposit_type == "No Deposit"),
    change_rate            = booking_changes / (lead_time + 1),
    no_deposit_x_lead      = no_deposit * lead_time,
    risky_booking          = no_deposit * prior_cancel_rate,
    no_deposit_no_requests = as.integer(no_deposit == 1 &
                                        total_of_special_requests == 0)
  ) |>
  step_impute_median(all_numeric_predictors()) |>
  step_impute_mode(all_nominal_predictors()) |>
  step_novel(all_nominal_predictors()) |>
  step_string2factor(all_nominal_predictors()) |>
  step_other(all_nominal_predictors(), threshold = 0.01) |>
  step_dummy(all_nominal_predictors()) |>          # one-hot encode for linear model
  step_normalize(all_numeric_predictors())

hotel_recipe_glm


# Gradient Boosting Trees Recipe
# GBM requires step_dummy
hotel_recipe_gbm <- recipe(is_canceled ~ ., data = hotel_bookings_train) |>
  step_rm(id) |>
  step_mutate(
    total_nights      = stays_in_weekend_nights + stays_in_week_nights,
    total_guests      = adults + children + babies,
    has_children      = as.integer(children > 0 | babies > 0),
    room_type_changed = as.integer(as.character(reserved_room_type) !=
                                   as.character(assigned_room_type)),
    prior_cancel_rate = previous_cancellations /
                        (previous_cancellations + previous_bookings_not_canceled + 1),
    no_deposit        = as.integer(deposit_type == "No Deposit"),
    change_rate       = booking_changes / (lead_time + 1),
    no_deposit_x_lead = no_deposit * lead_time,
    risky_booking     = no_deposit * prior_cancel_rate,
    no_deposit_no_requests = as.integer(no_deposit == 1 & total_of_special_requests == 0)
  ) |>
  step_impute_median(all_numeric_predictors()) |>
  step_impute_mode(all_nominal_predictors()) |>
  step_novel(all_nominal_predictors()) |>
  step_string2factor(all_nominal_predictors()) |>
  step_other(all_nominal_predictors(), threshold = 0.01) |>
  step_dummy(all_nominal_predictors())             # GBM needs numeric input; normalization not needed

hotel_recipe_gbm


# Random Forest Recipe
# ranger handles factors no dummy encoding needed
hotel_recipe <- recipe(is_canceled ~ ., data = hotel_bookings_train) |>
  step_rm(id) |>
  step_mutate(
    total_nights           = stays_in_weekend_nights + stays_in_week_nights,
    total_guests           = adults + children + babies,
    has_children           = as.integer(children > 0 | babies > 0),
    room_type_changed      = as.integer(as.character(reserved_room_type) !=
                                        as.character(assigned_room_type)),
    prior_cancel_rate      = previous_cancellations /
                             (previous_cancellations + previous_bookings_not_canceled + 1),
    no_deposit             = as.integer(deposit_type == "No Deposit"),
    change_rate            = booking_changes / (lead_time + 1),
    no_deposit_x_lead      = no_deposit * lead_time,
    risky_booking          = no_deposit * prior_cancel_rate,
    no_deposit_no_requests = as.integer(no_deposit == 1 &
                                        total_of_special_requests == 0)
  ) |>
  step_impute_median(all_numeric_predictors()) |>
  step_impute_mode(all_nominal_predictors()) |>
  step_novel(all_nominal_predictors()) |>
  step_string2factor(all_nominal_predictors()) |>
  step_other(all_nominal_predictors(), threshold = 0.01)

hotel_recipe


# KNN Recipe
hotel_recipe_knn <- recipe(is_canceled ~ ., data = hotel_bookings_train) |>
  step_rm(id) |>
  step_mutate(
    total_nights           = stays_in_weekend_nights + stays_in_week_nights,
    total_guests           = adults + children + babies,
    has_children           = as.integer(children > 0 | babies > 0),
    room_type_changed      = as.integer(as.character(reserved_room_type) !=
                                        as.character(assigned_room_type)),
    prior_cancel_rate      = previous_cancellations /
                             (previous_cancellations + previous_bookings_not_canceled + 1),
    no_deposit             = as.integer(deposit_type == "No Deposit"),
    change_rate            = booking_changes / (lead_time + 1),
    no_deposit_x_lead      = no_deposit * lead_time,
    risky_booking          = no_deposit * prior_cancel_rate,
    no_deposit_no_requests = as.integer(no_deposit == 1 & 
                                        total_of_special_requests == 0)
  ) |>
  step_impute_median(all_numeric_predictors()) |>
  step_impute_mode(all_nominal_predictors()) |>
  step_novel(all_nominal_predictors()) |>
  step_string2factor(all_nominal_predictors()) |>
  step_other(all_nominal_predictors(), threshold = 0.01) |>
  step_dummy(all_nominal_predictors()) |>
  step_normalize(all_numeric_predictors())         

hotel_recipe_knn


# Minimal Recipes (imputation only, no feature engineering)

# Minimal for RF + Single tree
hotel_recipe_minimal <- recipe(is_canceled ~ ., data = hotel_bookings_train) |>
  step_rm(id) |>
  step_impute_median(all_numeric_predictors()) |>
  step_impute_mode(all_nominal_predictors()) |>
  step_string2factor(all_nominal_predictors())

hotel_recipe_minimal


# Minimal for GBM
hotel_recipe_minimal_gbm <- recipe(is_canceled ~ ., data = hotel_bookings_train) |>
  step_rm(id) |>
  step_impute_median(all_numeric_predictors()) |>
  step_impute_mode(all_nominal_predictors()) |>
  step_string2factor(all_nominal_predictors()) |>
  step_dummy(all_nominal_predictors())

hotel_recipe_minimal_gbm


# Minimal for logistic regression
hotel_recipe_minimal_glm <- recipe(is_canceled ~ ., data = hotel_bookings_train) |>
  step_rm(id) |>
  step_impute_median(all_numeric_predictors()) |>
  step_impute_mode(all_nominal_predictors()) |>
  step_string2factor(all_nominal_predictors()) |>
  step_dummy(all_nominal_predictors()) |>
  step_normalize(all_numeric_predictors())

hotel_recipe_minimal_glm

```

### Table

| Technique \| Optimal ROC \| ROCSD \| Optimal Hyperparameter Values \|
| Single tree \| 0.8465006 \| 0.00811963 \| cp = 0.0005 \|
| Single tree (minimal FE - just imputation) \| 0.8313408 \| 0.0146011 \| cp = 0.001 \|
| Linear regression \| 0.865466 \| 0.01947199 \| alpha = 0.75, Lambda = 0.01128838 \|
| Linear regression (minimal FE - just imputation) \| 0.8533113 \| 0.02042113 \| alpha = 0.25, lambda = 0.0001 \|
| Boosted trees \| 0.8765175 \| 0.008613065 \| shrinkage = 0.01, interaction.depth = 15, n.minobsinnode = 1, n.trees = 400 \|
| Boosted trees (minimal FE - just imputation) \| 0.867072 \| 0.009904834 \| shrinkage = 0.01, interaction.depth = 15, n.minobsinnode = 1, n.trees = 600 \|
| Random forest \| 0.8758245 \| 0.006529793 \| mtry = 5, splitrule = gini, min.node.size = 3 \|
| Random forest (minimal FE - just imputation) \| 0.8647418 \| 0.008923356 \| mtry = 4, splitrule = gini, min.node.size = 1 \|

```{r}
### CV Specifications

cv_specs_hotel <- trainControl(method = "cv",
                               number = 5,
                               summaryFunction = twoClassSummary,
                               classProbs = TRUE,
                               savePredictions = "final")

```

```{r}
# Single Tree Model
# cp (complexity parameter) controls tree pruning so smaller cp = more complex tree.
tree_grid <- expand.grid(
  cp = c(0, 0.0001, 0.0005, 0.001, 0.005, 0.01, 0.05)
)

set.seed(208)
tree_hotel <- train(
  hotel_recipe_tree,
  data      = hotel_bookings_train,
  method    = "rpart",
  trControl = cv_specs_hotel,
  tuneGrid  = tree_grid,
  metric    = "ROC"
)

tree_hotel$results |>
  dplyr::filter(cp == tree_hotel$bestTune$cp)

```

```{r}
# Minimal Single Tree
set.seed(208)
tree_hotel_minimal <- train(
  hotel_recipe_minimal,
  data      = hotel_bookings_train,
  method    = "rpart",
  trControl = cv_specs_hotel,
  tuneGrid  = tree_grid,
  metric    = "ROC"
)
tree_hotel_minimal$results |>
  dplyr::filter(cp == tree_hotel_minimal$bestTune$cp)

```

```{r}
# Logistic Regression Model
# alpha controls ridge (0) vs lasso (1) penalty mix; lambda controls regularization strength.

glm_grid <- expand.grid(
  alpha  = c(0, 0.25, 0.5, 0.75, 1),
  lambda = 10^seq(-4, -1, length.out = 20)
)

set.seed(208)
glm_hotel <- train(
  hotel_recipe_glm,
  data      = hotel_bookings_train,
  method    = "glmnet",
  trControl = cv_specs_hotel,
  tuneGrid  = glm_grid,
  metric    = "ROC"
)

glm_hotel$results |>
  dplyr::filter(
    alpha  == glm_hotel$bestTune$alpha,
    lambda == glm_hotel$bestTune$lambda)

```

```{r}
# Minimal Logistic Regression
set.seed(208)
glm_hotel_minimal <- train(
  hotel_recipe_minimal_glm,
  data      = hotel_bookings_train,
  method    = "glmnet",
  trControl = cv_specs_hotel,
  tuneGrid  = glm_grid,
  metric    = "ROC"
)
glm_hotel_minimal$results |>
  dplyr::filter(
    alpha  == glm_hotel_minimal$bestTune$alpha,
    lambda == glm_hotel_minimal$bestTune$lambda
  )

```

```{r}
# Gradient Boosting Tree Model
# shrinkage (learning rate) is kept low (0.01) to allow gradual, stable learning.
# interaction.depth controls tree complexity; higher values capture more feature interactions.
# n.minobsinnode = 1 allows very small leaf nodes, maximizing fit on this dataset size.
gbm_grid <- expand.grid(
  n.trees           = c(400, 500, 600),
  interaction.depth = c(15, 17, 20),
  shrinkage         = 0.01,
  n.minobsinnode    = c(1,2,3)
)

set.seed(208)
gbm_hotel <- train(
  hotel_recipe_gbm,
  data       = hotel_bookings_train,
  method     = "gbm",
  trControl  = cv_specs_hotel,
  tuneGrid   = gbm_grid,
  metric     = "ROC",
  verbose    = FALSE
)

gbm_hotel$results |>
  tibble::rowid_to_column("Model") |>
  dplyr::filter(
    n.trees == gbm_hotel$bestTune$n.trees,
    interaction.depth == gbm_hotel$bestTune$interaction.depth,
    shrinkage == gbm_hotel$bestTune$shrinkage,
    n.minobsinnode == gbm_hotel$bestTune$n.minobsinnode
  )

# ROC/AUC
library(pROC)

gbm_roc <- roc(
  response = gbm_hotel$pred$obs,
  predictor = gbm_hotel$pred$yes,
  levels = c("no", "yes")
)

plot(gbm_roc, main = "CV ROC Curve - Boosted Trees")
auc(gbm_roc)

confusionMatrix(
  data = gbm_hotel$pred$pred,
  reference = gbm_hotel$pred$obs,
  positive = "yes"
)

# vip(gbm_hotel, num_features = 20)
gbm_hotel$bestTune

```

```{r}
# Minimal GBM
set.seed(208)
gbm_hotel_minimal <- train(
  hotel_recipe_minimal_gbm,
  data      = hotel_bookings_train,
  method    = "gbm",
  trControl = cv_specs_hotel,
  tuneGrid  = gbm_grid,
  metric    = "ROC",
  verbose   = FALSE
)
gbm_hotel_minimal$results |>
  dplyr::filter(
    n.trees           == gbm_hotel_minimal$bestTune$n.trees,
    interaction.depth == gbm_hotel_minimal$bestTune$interaction.depth,
    shrinkage         == gbm_hotel_minimal$bestTune$shrinkage,
    n.minobsinnode    == gbm_hotel_minimal$bestTune$n.minobsinnode
  )

```

```{r}
# 2nd version of Gradient Boosting Model
# PERFORMS WORSE THAN V1 so lower shrinkage with more trees did not improve ROC
set.seed(208)

gbm_grid_v2 <- expand.grid(
  n.trees = c(800, 1000, 1200),
  interaction.depth = c(15, 17, 20),
  shrinkage = c(0.005, 0.01),
  n.minobsinnode = c(1,2)
)

gbm_v2 <- train(
  hotel_recipe_gbm,         
  data      = hotel_bookings_train,
  method    = "gbm",
  trControl = cv_specs_hotel,
  tuneGrid  = gbm_grid_v2,  
  metric    = "ROC", 
  verbose = FALSE
)

gbm_v2$results |>
  tibble::rowid_to_column("Model") |>
  dplyr::filter(
    n.trees == gbm_v2$bestTune$n.trees,
    interaction.depth == gbm_v2$bestTune$interaction.depth,
    shrinkage == gbm_v2$bestTune$shrinkage,
    n.minobsinnode == gbm_v2$bestTune$n.minobsinnode)
    
```

```{r}
# Random Forest Model
set.seed(208)

rf_grid <- expand.grid(
  mtry = c(2, 3, 4, 5), 
  splitrule = "gini",
  min.node.size = c(1, 3, 5)
)

rf_hotel <- train(
  hotel_recipe,
  data = hotel_bookings_train,
  method = "ranger",
  trControl = cv_specs_hotel,
  tuneGrid = rf_grid,
  metric = "ROC"
)

rf_hotel$results |> 
  dplyr::filter(mtry == rf_hotel$bestTune$mtry,
                splitrule == rf_hotel$bestTune$splitrule,
                min.node.size == rf_hotel$bestTune$min.node.size)

```

```{r}
# Minimal RF
set.seed(208)
rf_hotel_minimal <- train(
  hotel_recipe_minimal,
  data      = hotel_bookings_train,
  method    = "ranger",
  trControl = cv_specs_hotel,
  tuneGrid  = rf_grid,
  metric    = "ROC"
)
rf_hotel_minimal$results |>
  dplyr::filter(
    mtry       == rf_hotel_minimal$bestTune$mtry,
    splitrule  == rf_hotel_minimal$bestTune$splitrule,
    min.node.size == rf_hotel_minimal$bestTune$min.node.size
  )

```

```{r}
# 2nd version of Random Forest Model
# PERFORMS WORSE THAN V1 so extending mtry beyond 5 did not improve ROC
set.seed(208)
rf_v2 <- train(
  hotel_recipe,
  data      = hotel_bookings_train,
  method    = "ranger",
  trControl = cv_specs_hotel,
  tuneGrid  = rf_hotel$bestTune,
  metric    = "ROC"
)

rf_v2$results |>
  dplyr::filter(
    mtry       == rf_v2$bestTune$mtry,
    splitrule  == rf_v2$bestTune$splitrule,
    min.node.size == rf_v2$bestTune$min.node.size
  )

```

```{r}
# KNN Model

set.seed(208)

knn_grid <- expand.grid(
  kmax     = c(201, 301, 401, 501),
  distance = 1,          # Manhattan distance — outperformed Euclidean in earlier tuning
  kernel   = "optimal"   # weights neighbors by distance
)

set.seed(208)
knn_hotel <- train(
  hotel_recipe_knn,
  data      = hotel_bookings_train,
  method    = "kknn",
  trControl = cv_specs_hotel,
  tuneGrid  = knn_grid,
  metric    = "ROC"
)

knn_hotel$results |>
  dplyr::filter(kmax == knn_hotel$bestTune$kmax)

```

\newpage

#### final predictions on the test set for submission to Kaggle

```{r}
# GBM test predictions — predicted probability of cancellation for each test booking
test_preds <- predict(gbm_hotel, newdata = hotel_bookings_test, type = "prob")
kaggle_submission_v1 <- data.frame(
  id          = hotel_bookings_test$id,
  is_canceled = test_preds$yes        # Kaggle expects probability of the positive class
)
write_csv(kaggle_submission_v1, "kaggle_submission_v1.csv")

# RF test predictions
test_preds_rf <- predict(rf_hotel, newdata = hotel_bookings_test, type = "prob")
kaggle_submission_v2 <- data.frame(
  id          = hotel_bookings_test$id,
  is_canceled = test_preds_rf$yes
)
write_csv(kaggle_submission_v2, "kaggle_submission_rf.csv")

# Ensemble — blend GBM and RF CV fold predictions
# Rows are joined by rowIndex to ensure predictions are aligned to the same observation
# before averaging; misaligned averaging would produce an AUC lower than either model alone
gbm_preds <- gbm_hotel$pred |>
  select(rowIndex, obs, gbm_yes = yes)

rf_preds <- rf_hotel$pred |>
  select(rowIndex, rf_yes = yes)

blended_preds <- gbm_preds |>
  inner_join(rf_preds, by = "rowIndex") |>
  mutate(blended = (gbm_yes + rf_yes) / 2)  # simple average blend

blended_roc <- roc(
  response  = blended_preds$obs,
  predictor = blended_preds$blended,
  levels    = c("no", "yes")
)
auc(blended_roc)   # ensemble AUC — should exceed both individual models

# Ensemble test submission — average predicted probabilities from both models
kaggle_submission_ensemble <- data.frame(
  id          = hotel_bookings_test$id,
  is_canceled = (test_preds$yes + test_preds_rf$yes) / 2
)
write_csv(kaggle_submission_ensemble, "kaggle_submission_ensemble.csv")


# Stacked Model — uses GBM, RF, and logistic regression predictions as inputs
# to a meta-learner (logistic regression), which learns the optimal blend weights
stack_train <- gbm_hotel$pred |>
  select(rowIndex, obs, gbm = yes) |>
  inner_join(rf_hotel$pred  |> select(rowIndex, rf  = yes), by = "rowIndex") |>
  inner_join(glm_hotel$pred |> select(rowIndex, glm = yes), by = "rowIndex") |>
  mutate(y = as.integer(obs == "yes"))

meta_model <- glm(y ~ gbm + rf + glm, data = stack_train, family = binomial)

# Stacked model AUC on CV folds
roc(stack_train$y, fitted(meta_model), levels = c(0,1)) |> auc()

# Seed-averaged ensemble — trains each model across multiple random seeds and
# averages test predictions to reduce variance from any single seed's randomness
seeds <- c(208, 42, 999, 1234, 5678)

gbm_seed_preds <- lapply(seeds, function(s) {
  set.seed(s)
  m <- train(
    hotel_recipe_gbm,
    data      = hotel_bookings_train,
    method    = "gbm",
    trControl = cv_specs_hotel,
    tuneGrid  = gbm_hotel$bestTune,   # fix hyperparameters, vary only the seed
    metric    = "ROC",
    verbose   = FALSE
  )
  predict(m, hotel_bookings_test, type = "prob")$yes
})
gbm_avg <- Reduce("+", gbm_seed_preds) / length(seeds)

rf_seed_preds <- lapply(seeds, function(s) {
  set.seed(s)
  m <- train(
    hotel_recipe,
    data      = hotel_bookings_train,
    method    = "ranger",
    trControl = cv_specs_hotel,
    tuneGrid  = rf_hotel$bestTune,    # fix hyperparameters, vary only the seed
    metric    = "ROC"
  )
  predict(m, hotel_bookings_test, type = "prob")$yes
})
rf_avg <- Reduce("+", rf_seed_preds) / length(seeds)

kaggle_seed_avg <- data.frame(
  id          = hotel_bookings_test$id,
  is_canceled = 0.5 * gbm_avg + 0.5 * rf_avg
)
write_csv(kaggle_seed_avg, "kaggle_seed_avg.csv")

```
